Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-10-04 doc/JFLA2022/paper.tex

\section{Modernizing the infrastructure}
\subsection{Documentation with Alectryon}
\begin{todo}
  Interest of Alectryon:
  \begin{itemize}
  \item Despite the frequent changes (improvements) in the Coq scripts, code inclusion and \textit{Coq}'s answers are up to date.
  \item \TODO[Clément]{A good part of the LaTeX support in Alectryon was developed for hydra-battles, and this was the first real application with a custom driver, too.}
  \item \TODO[Clément]{If you kept a trace of them, I would love to see a discussion of any errors that were caught by including the output systematically instead of copy-pasting it.}
  \item \TODO[Pierre]{the order of appearance of snippets in the book is independent from the module structure}
   \item \TODO[Pierre]{Slicing our proofs into snippets allowed to show only the main parts of our proofsleaved with text (Figure~\ref{fig:Ex42-schutte}).}

@@

   Figures~\ref{fig:Hprime} (code snippet), \ref{fig:Ex42E0} and \ref{fig:Ex42-schutte} have been generated with Alectryon.
\end{todo}

@@

>>>>

\section{Modernizing the infrastructure}

\subsection{Documentation with Alectryon}

The hydra-battles book is written in LaTeX, but it makes very frequent references (about once per page, 274 snippets over 281 pages) to parts of the Coq development, showing definitions (Fig.~\ref{fig:Hprime}), lemmas (Fig.~\ref{fig:Ex42E0}), proof scripts (Fig.~\ref{fig:Ex42-schutte}), and computation results (Fig.~\ref{fig:Pow-evalPow17LetIn}).

Originally, we copied snippets from Coq sources into LaTeX manually, and recorded and inserted the corresponding outputs manually as well.  This approach is common, but brittle: changes to Coq definitions or lemmas had to be reflected in the book's sources, and we found multiple instances where the book and the Coq development had diverged (see FIXME cite alectryon for a discussion of various approaches to documentation of Coq developments).

We solved this maintenance issue by moving to Alectryon, a tool to automatically record Coq proofs.  Instead of copy-pasting fragments into LaTeX, we now embed small LaTeX files automatically generated by Alectryon from our Coq development.  Our build system guarantees that these LaTeX snippets are always up-to-date and consistent with the code (and, by comparing these snippets across releases of Coq or versions of hydra-battles, we can easily spot unexpected changes).

The transition happened gradually, over a few months: for each snippet of Coq code that was in the book, we had to take the following steps:

\begin{enumerate}
\item Mark the snippet in the Coq sources (we use special comments \texttt{(* begin snippet \textit{name} *)} … \texttt{(* end snippet \textit{name} *)})
\item Configure output display, using special Alectryon annotations to configure what should be shown (only the inputs, inputs and outputs, some steps of the proof but not all, some proof states at key moments in a proof, etc.)
\item Replace the copy-pasted inputs and output in LaTeX with an \texttt{input} command.
\end{enumerate}

Importing snippets into a LaTeX document was not one of the original use cases of Alectryon, so we had to extend it: the original Alectryon did not have support for exporting to LaTeX, and it was towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended Alectryon to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom Alectryon \textit{driver}, a small (about 100 lines) Python program that leverages most of the Alectryon toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as Alectryon).

\begin{todo}
  Interest of Alectryon:
  \begin{itemize}
  \item \TODO[Pierre]{the order of appearance of snippets in the book is independent from the module structure}
   \item \TODO[Pierre]{Slicing our proofs into snippets allowed to show only the main parts of our proofsleaved with text (Figure~\ref{fig:Ex42-schutte}).}

@@

\end{todo}

@@

\begin{figure}
  \centering
  \inputsnippets{Pow/evalPow17LetIn}
  \caption{\label{fig:Pow-evalPow17LetIn}Automatically capturing the output of computations}
\end{figure}

### 2021-10-04 doc/JFLA2022/snippets/Schutte/Ex42c.tex

\begin{alectryon}
  % Generator: Alectryon
  \sep
  \begin{sentence}
    \begin{input}
      ~~\PY{n+nb}{assert}~\PY{o}{(}\PY{n}{HAP}\PY{o}{:}~\PY{n}{In}~\PY{n}{AP}~\PY{o}{(}\PY{n}{phi0}~\PY{l+m+mi}{2}\PY{o}{))}~\PY{n+nb+bp}{by}~\PY{n+nb}{apply}~\PY{n}{AP\PYZus{}phi0}\PY{o}{.}\nl
    \end{input}
  \end{sentence}
  \sep
  \begin{sentence}
    \begin{input}
      ~~\PY{n+nb}{elim}~~\PY{n}{HAP}\PY{o}{;}~\PY{n+nb}{intros}~~\PY{n}{\PYZus{}}~\PY{n}{H0}\PY{o}{;}~\PY{n+nb}{apply}~\PY{n}{H0}\PY{o}{;}~\PY{n+nb}{clear}~\PY{n}{H0}\PY{o}{.}
    \end{input}
    \sep
    \begin{output}
      \begin{goals}
        \begin{goal}
          \begin{hyps}
            \hyp{HAP}{\PY{n}{In}~\PY{n}{AP}~\PY{o}{(}\PY{n}{phi0}~\PY{o}{(}\PY{n}{F}~\PY{l+m+mi}{2}\PY{o}{))}}
          \end{hyps}
          \sep
          \infrule{}
          \sep
          \begin{conclusion}
            \PY{n}{omega}~\PY{o}{+}~\PY{n}{F}~\PY{l+m+mi}{42}~\PY{o}{\PYZlt{}}~\PY{n}{phi0}~\PY{o}{(}\PY{n}{F}~\PY{l+m+mi}{2}\PY{o}{)}
          \end{conclusion}
        \end{goal}
      \end{goals}
    \end{output}
  \end{sentence}
\end{alectryon}

>>>>

\begin{alectryon}
  % Generator: Alectryon
  \sep
  \begin{sentence}
    \begin{input}
      ~~\PY{n+nb}{assert}~\PY{o}{(}\PY{n}{Hlt}\PY{o}{:}~\PY{n}{omega}~\PY{o}{\PYZlt{}}~\PY{n}{phi0}~\PY{l+m+mi}{2}\PY{o}{)}~\PY{n+nb+bp}{by}~\nl
      ~~~~~~\PY{o}{(}\PY{n+nb}{rewrite}~\PY{n}{omega\PYZus{}eqn}\PY{o}{;}~\PY{n+nb}{apply}~\PY{n}{phi0\PYZus{}mono}\PY{o}{,}~\PY{n}{finite\PYZus{}mono}\PY{o}{;}\nl
      ~~~~~~~~\PY{n+nb}{auto~with}~\PY{n}{arith}\PY{o}{).}
    \end{input}
    \sep
    \begin{output}
      \begin{goals}
        \begin{goal}
          \begin{hyps}
            \hyp{HAP}{\PY{n}{In}~\PY{n}{AP}~\PY{o}{(}\PY{n}{phi0}~\PY{o}{(}\PY{n}{F}~\PY{l+m+mi}{2}\PY{o}{))}}
            \sep
            \hyp{Hlt}{\PY{n}{omega}~\PY{o}{\PYZlt{}}~\PY{n}{phi0}~\PY{o}{(}\PY{n}{F}~\PY{l+m+mi}{2}\PY{o}{)}}
          \end{hyps}
          \sep
          \infrule{}
          \sep
          \begin{conclusion}
            \PY{n}{omega}~\PY{o}{+}~\PY{n}{F}~\PY{l+m+mi}{42}~\PY{o}{\PYZlt{}}~\PY{n}{phi0}~\PY{o}{(}\PY{n}{F}~\PY{l+m+mi}{2}\PY{o}{)}
          \end{conclusion}
        \end{goal}
      \end{goals}
    \end{output}
  \end{sentence}
\end{alectryon}
